COFFEE: an Optimizing Compiler for Finite Element Local Assembly

### 2014-08-21 TACO-paper-v2/coffee.tex
Start writing the full application study

\subsection{The Helmholtz, Diffusion, and Burgers equations}
\label{sec:real-examples}
Our code transformations were evaluated in three real-world problems
based on the following PDEs
\begin{enumerate}
\item Helmholtz
\item Advection-Diffusion
\item Burgers
\end{enumerate}
The three chosen benchmarks are \emph{real-life kernels} and comprise
the core differential operators in some of the most frequently
encountered finite element problems in scientific computing. The
Helmholtz and Diffusion kernels are archetypal second order elliptic
operators. They are complete and unsimplified examples of the
operators occurring for diffusion and viscosity in fluids, and for
imposing pressure in compressible fluids. As such, they are both
extensively used in climate and ocean modeling. Very similar
operators, for which the same optimisations are expected to be equally
effective, apply to elasticity problems, which are at the base of
computational structural mechanics. The Burgers kernel is a typical
example of a first order hyperbolic conservation law, which occurs in
real applications whenever a quantity is transported by a fluid (the
momentum itself, in our case). We chose this particular kernel since
it applies to a vector-valued quantity, while the elliptic operators
apply to scalar quantities; this impacts the generated code, as
explained next. In this way, the operators we have selected are
characteristic of both the second and first order operators that
dominate in fluids and solids simulations, and a wide variety on local
assembly codes can be tested.

@@

These problems were studied varying both the shape of mesh elements
and the polynomial order $p$ of the method, whereas the element
family, Lagrange, is fixed. As might be expected, the larger the
element shape and $p$, the larger the iteration space. Triangles,
tetrahedron, and prisms were tested as element shape. For instance, in
the case of Helmholtz with $p=1$, the size of the \texttt{j} and
\texttt{k} loops for the three element shapes is, respectively, $3$,
$4$, and $6$. Moving to bigger shapes has the effect of increasing the
number of basis function arrays, since, intuitively, the behavior of
the equation has now to be approximated also along a third axis. On
the other hand, the polynomial order affects only the problem size
(the three loops \texttt{i}, \texttt{j}, and \texttt{k}, and, as a
consequence, the size of $X$ and $Y$ arrays). A range of polynomial
orders from $p=1$ to $p=4$ were tested; higher polynomial orders are
excluded from the study because of current Firedrake limitations. In
all these cases, the size of the element matrix rarely exceeds
30$\times$30, with a peak of 105$\times$105 in Burgers with prisms and
$p=4$.

\subsection{Loop permutation}
In the following, only results for the loop order \texttt{ijk} are
shown. For the considerations exposed in
Section~\ref{sec:pyop2-compiler}, loop interchanges having an inner
loop along \texttt{i} caused slow downs; also, interchanging
\texttt{j} and \texttt{k} loops while keeping \texttt{i} as outermost
loop did not provide any benefits.

\subsection{Impact of Generalized Loop-invariant Code Motion}
\label{licm-impact}

@@

\subsection{Impact of Padding and Data Alignment}

@@

\subsection{Impact of Vector-register Tiling}
\label{sec:perf-eval-opvect}
In this section, we evaluate the impact of vector-register tiling. We
compare two versions: the baseline, \emph{licm-ap}; and
vector-register tiling on top of \emph{licm-ap}, which in the
following is referred to simply as \emph{op-vect}.

@@

\subsection{Impact of Expression Splitting}
\label{sec:perf-results-split}
Expression splitting relieves the register pressure when the element
matrix evaluation needs to read from a large number of basis function
arrays. As detailed in Section~\ref{sec:expr-split}, in the price to
pay for this optimazion, there are an increased number of accesses to
the element matrix and, potentially, redundant computation. Similarly
to the analysis of vector-register tiling, we compare two versions:
the baseline, \emph{licm-ap}; and expression splitting on top of
\emph{licm-ap}, which, for simplicity, in the following is referred to
as \emph{split}.

@@

\subsection{Comparison with FEniCS Form Compiler's built-in Optimizations}
\label{sec:perf-results-fenics}
We have modified the FEniCS Form Compiler (FFC) to return an abstract
syntax tree representation of a local assembly kernel, rather than
plain C++ code, so as to enable code transformations in
COFFEE. Besides Firedrake, FFC is used in the FEniCS
project~\cite{Fenics}. In FEniCS, FFC can apply its own model-driven
optimizations to local assembly kernels~\cite{quadrature1}, which
mainly consist of loop-invariant code motion and elimination, at code
generation time, of floating point operations involving zero-valued
entries in basis function arrays.

@@

\subsection{Comparison with hand-made BLAS-based implementations}
\label{sec:perf-eval-blas}
For the Helmholtz problem on a tetrahedral mesh, manual
implementations based on \texttt{Intel MKL BLAS} were tested on the
Sandy Bridge. This particular kernel can be easily reduced to a
sequence of four matrix-matrix multiplies that can be computed via
calls to BLAS \texttt{dgemm}. In the case of $p=4$, where the element
matrix is of size 35$\times$35, the computation was almost twice
slower than the case in which \emph{licm-ap} was used, with the slow
down being even worse for smaller problem sizes. These experiments
suggest that the question regarding to what extent linear algebra
libraries can improve performance cannot be trivially answered. This
is due to a combination of issues: the potential loss in data
locality, as exposed in Section~\ref{sec:expr-split}, the actual
effectiveness of these libraries when the arrays are relatively small,
and the problem inherent to assembly kernels concerning extraction of
matrix-matrix multiplies from static analysis of the kernel's code. A
comprehensive study of these aspects will be addressed in further
work.

>>>>

We present two studies. Firstly, in Section~\ref{sec:analysis-code-transf}, we
analyze the impact of the proposed code transformations, and their interplay,
in three real-world equations. We quantify the speed ups achievable at
the level of the local assembly kernel, i.e. by excluding from the measurements
the other sections of the computation. Then, in Section~\ref{sec:full-app}, we
provide a full-application investigation. Here, we use another equation (so as to 
bring the total number of examined problems to four) to demonstrate that COFFEE's
optimizations actually allow a significant reduction in the overall computation run-time 
to be achieved, which practically motivates our research.

\subsection{Impact of Code Transformations}
\label{sec:analysis-code-transf}

\subsubsection{The Helmholtz, Diffusion, and Burgers equations}
\label{sec:real-examples}
Our code transformations were evaluated in three real-world problems
based on the following PDEs
\begin{enumerate}
\item Helmholtz
\item Advection-Diffusion
\item Burgers
\end{enumerate}
The three chosen benchmarks are \emph{real-life kernels} and comprise
the core differential operators in some of the most frequently
encountered finite element problems in scientific computing. The
Helmholtz and Diffusion kernels are archetypal second order elliptic
operators. They are complete and unsimplified examples of the
operators occurring for diffusion and viscosity in fluids, and for
imposing pressure in compressible fluids. As such, they are both
extensively used in climate and ocean modeling. Very similar
operators, for which the same optimisations are expected to be equally
effective, apply to elasticity problems, which are at the base of
computational structural mechanics. The Burgers kernel is a typical
example of a first order hyperbolic conservation law, which occurs in
real applications whenever a quantity is transported by a fluid (the
momentum itself, in our case). We chose this particular kernel since
it applies to a vector-valued quantity, while the elliptic operators
apply to scalar quantities; this impacts the generated code, as
explained next. In this way, the operators we have selected are
characteristic of both the second and first order operators that
dominate in fluids and solids simulations, and a wide variety on local
assembly codes can be tested.

@@

These problems were studied varying both the shape of mesh elements
and the polynomial order $p$ of the method, whereas the element
family, Lagrange, is fixed. As might be expected, the larger the
element shape and $p$, the larger the iteration space. Triangles,
tetrahedron, and prisms were tested as element shape. For instance, in
the case of Helmholtz with $p=1$, the size of the \texttt{j} and
\texttt{k} loops for the three element shapes is, respectively, $3$,
$4$, and $6$. Moving to bigger shapes has the effect of increasing the
number of basis function arrays, since, intuitively, the behaviour of
the equation has now to be approximated also along a third axis. On
the other hand, the polynomial order affects only the problem size
(the three loops \texttt{i}, \texttt{j}, and \texttt{k}, and, as a
consequence, the size of $X$ and $Y$ arrays). A range of polynomial
orders from $p=1$ to $p=4$ were tested; higher polynomial orders are
excluded from the study because of current Firedrake limitations. In
all these cases, the size of the element matrix rarely exceeds
30$\times$30, with a peak of 105$\times$105 in Burgers with prisms and
$p=4$.

\subsubsection{Loop interchange}
In the following, only results for the loop order \texttt{ijk} are
shown. For the considerations exposed in
Section~\ref{sec:pyop2-compiler}, loop interchanges having an inner
loop along \texttt{i} caused slow downs; also, interchanging
\texttt{j} and \texttt{k} loops while keeping \texttt{i} as outermost
loop did not provide any benefits.

\subsubsection{Impact of Generalized Loop-invariant Code Motion}
\label{licm-impact}

@@

\subsubsection{Impact of Padding and Data Alignment}

@@

\subsubsection{Impact of Vector-register Tiling}
\label{sec:perf-eval-opvect}
In this section, we evaluate the impact of vector-register tiling. We
compare two versions: the baseline, \emph{licm-ap}; and
vector-register tiling on top of \emph{licm-ap}, which in the
following is referred to simply as \emph{op-vect}.

@@

\subsubsection{Impact of Expression Splitting}
\label{sec:perf-results-split}
Expression splitting relieves the register pressure when the element
matrix evaluation needs to read from a large number of basis function
arrays. As detailed in Section~\ref{sec:expr-split}, in the price to
pay for this optimazion, there are an increased number of accesses to
the element matrix and, potentially, redundant computation. Similarly
to the analysis of vector-register tiling, we compare two versions:
the baseline, \emph{licm-ap}; and expression splitting on top of
\emph{licm-ap}, which, for simplicity, in the following is referred to
as \emph{split}.

@@

\subsubsection{Comparison with FEniCS Form Compiler's built-in Optimizations}
\label{sec:perf-results-fenics}
We have modified the FEniCS Form Compiler (FFC) to return an abstract
syntax tree representation of a local assembly kernel, rather than
plain C++ code, so as to enable code transformations in
COFFEE. Besides Firedrake, FFC is used in the FEniCS
project~\cite{Fenics}. In FEniCS, FFC can apply its own model-driven
optimizations to local assembly kernels~\cite{quadrature1}, which
mainly consist of loop-invariant code motion and elimination, at code
generation time, of floating point operations involving zero-valued
entries in basis function arrays.

@@

\subsubsection{Comparison with hand-made BLAS-based implementations}
\label{sec:perf-eval-blas}
For the Helmholtz problem on a tetrahedral mesh, manual
implementations based on \texttt{Intel MKL BLAS} were tested on the
Sandy Bridge. This particular kernel can be easily reduced to a
sequence of four matrix-matrix multiplies that can be computed via
calls to BLAS \texttt{dgemm}. In the case of $p=4$, where the element
matrix is of size 35$\times$35, the computation was almost twice
slower than the case in which \emph{licm-ap} was used, with the slow
down being even worse for smaller problem sizes. These experiments
suggest that the question regarding to what extent linear algebra
libraries can improve performance cannot be trivially answered. This
is due to a combination of issues: the potential loss in data
locality, as exposed in Section~\ref{sec:expr-split}, the actual
effectiveness of these libraries when the arrays are relatively small,
and the problem inherent to assembly kernels concerning extraction of
matrix-matrix multiplies from static analysis of the kernel's code. A
comprehensive study of these aspects will be addressed in further
work.

\subsection{Full-application Study}
\label{sec:full-app}
In this section, we investigate the performance gain for an entire finite element computation
developed in Firedrake, a static linear elasticity problem. The source code is available at~\cite{elasticity-code}. The application simulates ... .

The execution time of a (non-explicit) finite element method is dominated by two factors: assembly and solve. Once computed, the local element matrices and vectors are respectively inserted in a ``global'' sparse matrix and a global vector; these form a linear system, which is usually solved with an iterative method (the ``solve'' phase). The amount of time spent in assembly and solve varies from problem to problem. As reiterated in the available literature, for example in~\cite{quadrature1}, the computational cost of local assembly grows with the complexity of the partial differential equation (e.g. because of larger loops, more articulated expressions), while the solve, at worst, exhibits a minimal increase (what impacts the solve are, for instance, polynomial order and mesh discretization). The complexity of an equation depends on several factors, including the number of derivatives and coefficient functions (i.e. known functions that are prescribed with the finite element specification). We study two versions of our static elasticity problem: 1) with one coefficient function ($f=1$) and 2) with two coefficient functions ($f=2$). More coefficient functions are also plausible; however, as we will see, the assembly process starts being particularly expensive already in case 1).

The application runs on a tetrahedral mesh composed of 196608 cells. In the two versions of the elasticity problem, we examine the cases of polynomial order $p=1$ and $p=2$. To run our experiments, we use a single core of the Sandy Bridge architecture described in Section~\ref{sec:perf-results}.

In Figure~\ref{fig:full-app-res}, we show the execution times for the four test cases, without and with optimizations enabled. There are four groups of two bars, each group corresponding to a specific test case; for example, the name $f_1p_1$ below the first group of bars refers to the case in which the elasticity problem has 1 coefficient function and has polynomial order 1. A stacked bar captures the time spent in the assembly phase (\texttt{assembly}), to solve the linear system (\texttt{solve}), and in the other various parts of the program (\texttt{other} - for example, setup of the problem and initialization of the coefficient functions). The cost of the insertion of the computed local element matrices (and vectors) in the global matrix (vector) is incorporated in \texttt{assembly}.

We first notice that, in the scenario $f_1p_1$, the assembly is dominated by matrix insertion: despite the application of generalized loop-invariant code motion, data alignment and padding, there is only a minimal performance gain. This changes instantly moving to $p=2$ and $f=2$. In these scenarios, not only the cost of \texttt{assembly} becomes larger with respect to \texttt{solve}, but also does it make insertion cost negligible. The transformations automatically applied by COFFEE successfully decrease the impact of \texttt{assembly} over \texttt{solve}: this was achieved by a combined application of generalized loop-invariant code motion, alignment and padding, and expression splitting. Interesting are two facts. First, generalized loop-invariant code motion was particularly intensive: ... . Second, ... .

%We conclude noting that in explicit finite element methods you would have more local assembly , more often, no solver
